---
title: "ACER: Analysis of CRISPR Essentiality in R"
author: 
  - name: Elizabeth Hutton
    affiliation: 1
  
address:
  - code: 1
    address: Cold Spring Harbor Laboratory; Syosset, NY, USA; \url{orcid.org/0000-0003-3422-5278}

lead_author_surname: Hutton
doi: "https://github.com/CshlSiepelLab/ACE"
output: 
  pinp::pinp
  
keep_tex: false
vignette: >
  %\VignetteIndexEntry{ACER: Analysis of CRISPR Essentaility in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignettePackage{ACER}
  
acknowledgements: |
  This tool was designed in collaboration with Prof. Adam Siepel and Prof. 
  Chris Vakoc at Cold Spring Harbor Laboratory.
keywords:
  - gene essentiality
  - computational biology
  - CRISPR
  - differential essentiality
  - gene addiction

bibliography: ace
watermark: false
footer_contents: "ACER Vignette"
link-citations: true
one_column: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Tool Applications
  ACE (Analysis of CRISPR Essentiality) was designed to test for differential essentiality
between sets of samples from a CRISPR knockout screen. Please see the ACE paper
for a detailed discussion of methods. [@Hutton2020]

## Input Data
  A minimum of two data files must be provided for ACE: 
  
* `countFile`: a single text file with unique sgRNA identifiers in the first column and
gene names in the second column.  Subsequent columns must contain either depleted read counts
(if a master library sequencing is provided),
or alternating initial and depleted read counts.
* `negCtrlFile`: a text file of the gene names to use as negative controls.

Other files necessary for analysis may include:

* `masterFiles`: a vector of names of files containing sequenced master libraries.
* `sampleMasterInfoFile`: Text file containing a column of master library file names,
and a column with the corresponding sample name derived from that master library
(same as the `countFile` headers).
* `sampleInfoFile`: Required for differential essentiality prediction, this file
contains one column listing sample names and a second column containing sample
annotations to use in test partitioning (for instance, 'KRAS_WT' vs. 'KRAS_Mut').
* `guideCovarFile`: File used to estimate guide efficiency (optional).

```{r setup}
library(ACER)
```
## Tool Applications
ACE (Analysis of CRISPR Essentiality) was designed to test for differential essentiality
between sets of samples from a CRISPR knockout screen. Please see the ACE paper
for a detailed discussion of methods. [@Hutton2020]

## Input Data
A minimum of two data files must be provided for ACE: 
`countFile`: a single text file with unique sgRNA identifiers in the first column and
gene names in the second column.  Subsequent columns must contain either depleted read counts
(if a master library sequencing is provided),
or alternating initial and depleted read counts.
`negCtrlFile`: a text file of the gene names to use as negative controls.

Other files necessary for analysis may include:
`masterFiles`: a vector of names of files containing sequenced master libraries.
`sampleMasterInfoFile`: Text file containing a column of master library file names,
and a column with the corresponding sample name derived from that master library
(the same sample name used as a header in the `countFile`).
`sampleInfoFile`: Required for differential essentiality prediction, this file
contains one column listing sample names and a second column containing sample
annotations to use in test partitioning (for instance, 'KRAS_WT' vs. 'KRAS_Mut').
`guideCovarFile`: File used to estimate guide efficiency (optional).


```{r sample inputs}
head(fread(system.file('extdata','countData.csv', package='ACER')))
```
## Basic Analysis
The basic workflow for using ACE is three commands.  First, all input data files
are loaded into an R object (`DataObj`) and checked for problems.  Next, this raw data is
preprocessed according to the desired analysis to create a collection of parameters,
forming a `ModelObj`.  Both the raw data and the estimated parameters are then
provided to the main function, `optimizeModelParameters`, which uses expectation
maximization to infer gene essentiality, guide efficiency, and sample effects.
```{r basic}
newDataObj <- DataObj$new(masterFiles = system.file('extdata','masterLibraryCounts.csv', package='ACER'),
                          countFile = system.file('extdata','countData.csv', package='ACER'),
                          negCtrlFile = system.file('extdata','negCtrlGenes.txt', package='ACER'), 
                          sampleInfoFile=system.file('extdata','sampleAnnotations.txt', package='ACER'),
                          hasInitSeq = T)
newModelObj <- ModelObj$new(user_DataObj = newDataObj,
                            use_neg_ctrl=T,
                            test_samples='test',
                            use_master_library = T)
newResultsObj <- optimizeModelParameters(user_DataObj = newDataObj,
                                         user_ModelObj = newModelObj)
writeResObj(newResultsObj)
```
                                          
## Differential Essentiality
To obtain estimates of differential essentiality estimated by sample subtype,
as opposed to global essentiality across all samples, the file
`sampleInfoFile` must be provided with sample annotation information 
to the original `DataObj`. Essentiality will be compared between samples with
and without the annotation provided in the `ModelObj`'s `test_samples` parameter.

